---
title: "R Workshop 2 (Term 2)"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

# Introduction

Hello everyone! Here is a chance to explore some of the code and concepts covered in your QS903 lectures.

Specifically, we cover:

* The curve function
* The %*% operator
* The proportional odds model (for ordered categorical outcomes)
* The multinomial model (for unordered categorical outcomes)

You have also covered the multi-level models. These are cool and exciting! However, we are limited to two hours. So this workshop focuses on the models you may use in your first assignments.

# The curve function

Imagine you want to plot a function. One way to do this is to define a function,

```{r myFunction}
myInverseLogit <- function(x){
  exp(x)/(1+exp(x)) # this is the inverse logit function, btw
}
```

generate some value from the function

```{r functionPredictions}
x <- seq(from = -5, to = 5, by = 0.1)
y <- myInverseLogit(x)
```

and then plot a line using those values

```{r plotFunctionPredictions}
plot(x, y, type = 'l', ylab = 'myInverseLogit(x)') # type is line 'l'
```

The curve function combines generating predictions and plotting a line into one command.

```{r curve}
curve(expr = myInverseLogit, from = -5, to = 5, n = 100)
```

Nice. You may remember using the inverse logit function for the logit model. The invlogit function from the arm package you gives us the same predictions.

```{r arm_invlogit}
require(arm)
curve(expr = invlogit, from = -5, to = 5, n = 100)
```

An interesting feature of curve is that it is easy to overlay a curve onto a base R plot. 

```{r addCurve}
x <- seq(from = -5, to = 5, by = 1)
plot(x = x, y = invlogit(x))
curve(expr = invlogit, from = -5, to = 5, n = 100, add = TRUE, col = 'pink') # note add is TRUE
```

Without the add parameter curve creates a new plot.

```{r oopsCurve}
x <- seq(from = -5, to = 5, by = 1)
plot(x = x, y = invlogit(x))
curve(expr = invlogit, from = -5, to = 5, n = 100, col = 'purple')
```

# The %*% operator

This is the maxtix multiplication operator. Let us take a [simple example](https://www.mathsisfun.com/algebra/matrix-multiplying.html).

We will define two matrices from the example in the link above.

```{r matrices}
a <- matrix(data = c(1,4,2,5,3,6), nrow = 2, ncol = 3)
b <- matrix(data = c(7,9,11,8,10,12), nrow = 3, ncol = 2)
print(a)
print(b)
```

and then use the matrix multiplication operator to calculate the dot product.

```{r dotProduct}
print(a %*% b)
```

In the lecture notes, the inverse logit was used to generate predictions which were drawn onto a plot using the curve function.

# The proportional odds model

Let us walk through the multinomial example given in the lecture for week 3.

Load in the data.

```{r lecture_propOdds_data}
library(car)
data(Womenlf) # load in the womenlf data
head(Womenlf) # top of data
str(Womenlf) # structure of the data
```

The structure shows us there are 3 factors and 1 integer. For the proportional odds model we want to set our outcome as an ordered factor. Recoding is simply adding some meta data to a series of strings (e.g., category labels, order of categories, etc.).

Let us recode the partic variable. To illustrate how recoding works we will turn partic into a string, string to factor, and ordered factor.

```{r recoding}

a <- as.character(Womenlf$partic)
b <- as.factor(Womenlf$partic)
c <- ordered(Womenlf$partic, 
             levels = c('not.work', 'parttime', 'fulltime'))
Womenlf$status <- ordered(Womenlf$partic, 
             levels = c('not.work', 'parttime', 'fulltime'))
class(a)
class(b)
class(c)
```

```{r recoding_structure}
str(a)
str(b)
str(c)
```

The ordered factor is entered as an outcome in a proportional odds model. We want to examine the impact of the presence of children and husbands income predicts the work status of women.

```{r polr}
library(MASS)
fit.pom <- polr(status ~ hincome + children,
               data=Womenlf)
summary(fit.pom)
```

To interpret these results we calculate predicted probabilities. We have two cuts: not.work|parttime and parttime|fulltime. Using the coefficients for these cuts we can calculate the probabilities.

For the first cut

$$P(not working) = 1 - logit^{-1}(-0.5 \cdot hincome - 2 \cdot childrenpresent + 1.9)$$

which in R with no children and a husband income of 10 is

```{r basicProb}
print('The coefficients of the model are')
coef(fit.pom)

print('The sum of multiplying each predictor value by the corresponding coefficient is')
cbind(10,0)%*%coef(fit.pom)

print('Probability of not working with no children and a husbands income of 10')
1 - invlogit(cbind(10,0)%*%coef(fit.pom) + 1.9)

# note we multiply the values of children present (0) and husbands income (10) 
# by the corresponding coefficients using the matrix multiplication

```

You are given a neat function to calculate these values for each of the three categories for the corresponding values of hincome and child.

```{r firstCuttoff}
require(arm)
p.pom <- function(X, fit){
  b <- coef(fit)
  cuts <- fit$zeta
  p1 <- 1 - invlogit(X%*%b - cuts[1])
  p2 <- invlogit(X%*%b - cuts[1]) - 
    invlogit(X%*%b - cuts[2])
  p3 <- invlogit(X%*%b - cuts[2])
  cbind(p1, p2, p3)
}
```

For example, the probablity predicted by the model when a husbands income is 10 and there is no child is

```{r pomExample}
p.pom(c(10,0), fit.pom)
```

It is more likely that a women will work if the husband has an income of 10 and there is no child. Whereas if the husband is on a high income and there is a child then

```{r secondPomExample}
p.pom(c(40,1), fit.pom)
```

You were shown in the lecture how to plot the probability curves for each category.

```{r plotChildrenAbsent}
plot(c(1,45), c(0,1), type="n", main="Children absent",
     xlab="Husband's income", ylab="Pred. Probability")
curve(p.pom(cbind(x,0), fit.pom)[,1], add=TRUE, lty=1)
curve(p.pom(cbind(x,0), fit.pom)[,2], add=TRUE, lty=2)
curve(p.pom(cbind(x,0), fit.pom)[,3], add=TRUE, lty=3)
# From and To not given so over plot range
legend("topright", lty=1:3, 
       c("not working", "part-time","full-time"), 
       bty="n")
```

```{r plotChildrenPresent}
plot(c(1,45), c(0,1), type="n", main="Children present",
xlab="Husband's income", ylab="Pred. Probability")
curve(p.pom(cbind(x,1), fit.pom)[,1], add=TRUE, lty=1)
curve(p.pom(cbind(x,1), fit.pom)[,2], add=TRUE, lty=2)
curve(p.pom(cbind(x,1), fit.pom)[,3], add=TRUE, lty=3)
legend("topleft", lty=1:3, 
       c("not working", "part-time","full-time"), 
       bty="n")
```

## Diagnostics

How can we tell is our model is doing a good job? Or is the data is probably unordered?

Well, we can look at the residuals. The diffence betweeen the expected value and the observed value.

```{r residuals}
y = as.numeric(Womenlf$status)
expected.value = fitted(fit.pom)%*%c(1,2,3)
residual = y - expected.value
```

The plot used in the lecture is a binned plot. You can look at the help functions for this using ?binnedplot. 

```{r binnedPlot}
binnedplot(expected.value, residual, nclass=7)
?binnedplot
```

How does that look to you? Does it look quite reasonable?

If we look at this for children.

```{r residualChildren}
binnedplot(as.numeric(Womenlf$children), residual, xlab="Children")
```

Thoughts?

How about for husbands income?

```{r residualIncome}
binnedplot(as.numeric(Womenlf$hincome), residual,  nclass = 7, xlab = "Husbands income")
```

It seems there is a falling pattern of the residuals. Perhaps status is unordered!

Time to panic? Nah, we have a model for unordered categorical data.

# Multinomial Logit Model

The model can fit categorical outcomes without an order.

```{r fitMultinomial}
require(nnet)
fit.mlm <- multinom(status ~ hincome + children,
                    data = Womenlf)
summary(fit.mlm)
```

And to calculate the probability of the second category (full-time) is,

$$P(\textit{part-time}) = \frac{\exp(X_{i}\beta_{2})}{1 + \exp(X_{i} \beta_{2}) + \exp(X_{i} \beta_{3})}$$

Which where there is no child and the husbands income is 10 is,

```{r multinomialPrediction}
X <- cbind(1, 10, 0)
b <- coef(fit.mlm)
exp(X%*%b[1,]) /
  (1 + exp(X%*%b[1,]) + exp(X%*%b[2,]))
```

Andreas provides you with a nifty function to calculate the probabilities of all the categories.

```{r allTheCategories}

p.mlm <- function(X, fit){
  b <- coef(fit)
  p2 <- exp(X%*%b[1,]) /
    (1 + exp(X%*%b[1,]) + exp(X%*%b[2,]))
  p3 <- exp(X%*%b[2,]) /
    (1 + exp(X%*%b[1,]) + exp(X%*%b[2,]))
  p1 <- 1 - p2 - p3
  cbind(p1, p2, p3)
}

```

So, the 

```{r multinomExample}
p.mlm(c(1,10,0), fit.mlm)
```

When there is no child and the husbands income is 10 then it is most likely a women is working full time.

As before, we can plot the predicted probablities for when there is no child

```{r mlmNoChild}
plot(c(1,45), c(0,1), type="n", main="Children absent",
xlab="Husband's income", ylab="Pred. probability")
curve(p.mlm(cbind(1,x,0), fit.mlm)[,1], add=TRUE, lty=1)
curve(p.mlm(cbind(1,x,0), fit.mlm)[,2], add=TRUE, lty=2)
curve(p.mlm(cbind(1,x,0), fit.mlm)[,3], add=TRUE, lty=3)
legend("topright", lty=1:3, c("not working","part-time",
"full-time"), bty="n")
```

and when there is a child

```{r mlmChild}
plot(c(1,45), c(0,1), type="n", main="Children present",
xlab="Husband's income", ylab="Pred. probability")
curve(p.mlm(cbind(1,x,1), fit.mlm)[,1], add=TRUE, lty=1)
curve(p.mlm(cbind(1,x,1), fit.mlm)[,2], add=TRUE, lty=2)
curve(p.mlm(cbind(1,x,1), fit.mlm)[,3], add=TRUE, lty=3)
legend("topleft", lty=1:3, c("not working","part-time",
"full-time"), bty="n")
```

We need to look at the residuals per category.

```{r per_category}
# predicted probabilities
p.ful = fitted(fit.mlm)[,"fulltime"]
p.par = fitted(fit.mlm)[,"parttime"]
p.not = fitted(fit.mlm)[,"not.work"]
# observed outcomes as binary variables
y.ful = ifelse(Womenlf$status=="fulltime", 1, 0)
y.par = ifelse(Womenlf$status=="parttime", 1, 0)
y.not = ifelse(Womenlf$status=="not.work", 1, 0)
# residuals = observed - predicted
r.ful = y.ful - p.ful
r.par = y.par - p.par
r.not = y.not - p.not
```

We can examine these in detail if needed. The ifelse statements return a 1 is the first argument is TRUE and 0 if it is FALSE.

Which we can then plot.

```{r binplot1}
binnedplot(p.ful, r.ful, nclass=7, xlab = 'Predicted probability', main = 'Full-time')
```

```{r binplot2}
binnedplot(p.par, r.par, nclass=7, xlab = 'Predicted probability', main = 'Part-time')
```

```{r binplot3}
binnedplot(p.not, r.not, nclass=7, xlab = 'Predicted probability', main = 'Not working')
```

And once again by a predictor (husband's income)

```{r mlmPredictorBin}
par(mfrow = c(1,3), mar = c(3,4,2,1), mgp = c(2,.7,0), cex=1.2)
binnedplot(Womenlf$hincome, r.ful, nclass = 7, xlab = 'Husbands income', main = 'Full-time')
binnedplot(Womenlf$hincome, r.par, nclass = 7, xlab = 'Husbands income', main = 'Part-time')
binnedplot(Womenlf$hincome, r.not, nclass = 7, xlab = 'Husbands income', main = 'Not working')
```

Does the multinomial model do better? Does a model with unordered categories outperform a model with ordered categories?

Andreas uses an informal test of the proportional odds assumption.

```{r propOddsAssum}
c(deviance(fit.pom), deviance(fit.mlm), fit.pom$edf, fit.mlm$edf)
```

A reduction in deviance of 19 with only 2 more parameters is not too bad. This is only an informal test though. Generally the models predictions are similiar, which is also reassuring.

# Things to do

We have reconsidered the code presented by Andreas in your week 3 lecture. Now you are going to use this code to understand new data sets. There are 3 data sets:

* Student survey
* Marvel comic characters
* DC comic characters

## Student survey

In the MASS package is a data set called 'survey'. It is a survey of 271 University students.

```{r surveyLoad}
require(MASS)
data('survey')
str(survey)

# you may need to remove NA values using is.na
```

Your job is to address the question:

* Do age and sex contribute to a students probability of smoking?

## Marvel and DC

The blog fivethirtyeight published an analysis of [comic book characters and gender](https://github.com/fivethirtyeight/data/tree/master/comic-characters). This is quite a fun data set.

```{r comicBookCharacters} 

dc <- read.csv(file = 'https://github.com/fivethirtyeight/data/blob/master/comic-characters/dc-wikia-data.csv?raw=true',
                    stringsAsFactors = FALSE)

marvel <- read.csv(file = 'https://github.com/fivethirtyeight/data/blob/master/comic-characters/marvel-wikia-data.csv?raw=true',
                    stringsAsFactors = FALSE)

print('The DC data set')
str(dc)

print('The marvel data set')
str(marvel)
# you will need to recode data using factor and ordered
```

For each data set, can you find out:

* How does the sex (gender) and year of introduction contribute to the alignment of comic book characters?
* Does this differ between DC and Marvel?

Good luck!